---
title: "Visualize GatingSet with ggcyto"
output:
  html_document:
    fig_height: 3
    fig_width: 4
    keep_md: yes
vignette: >    
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Visualize GatingSet with ggcyto}        
---

```{r, echo=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


```{r echo=FALSE}
library(ggcyto)
dataDir <- system.file("extdata",package="flowWorkspaceData")
gs <- load_gs(list.files(dataDir, pattern = "gs_manual",full = TRUE))
```

By specifying the dimensions through `aes` and selecting the cell population through `subset`, `ggcyto` can easily visualize the gated data stored in `GatingSet`.
```{r}
p <- ggcyto(gs, aes(x = CD4, y = CD8), subset = "CD3+") 
# 2d plot 
p <- p + geom_hex(bins = 64)
p
```

We can use the instrument range to automatically filter out these outlier cell events
```{r}
p + ggcyto_par_set(limits = "instrument")
```

Or by setting limits manually
```{r}
myPars <- ggcyto_par_set(limits = list(x = c(0,3.5e3), y = c(-10, 4.1e3)))
p <- p + myPars# or xlim(0,3.5e3) + ylim(-10, 4e3) 
p
```

To check what kind of visualization parameters can be changed through `ggcyto_par_set`, simply print the default settings
```{r}
ggcyto_par_default()
```

To plot a gate, simply pass the gate name to the `geom_gate` layer
```{r}
p + geom_gate("CD4")
```


More than one gate can be added as long as they share the same parent and dimensions
```{r}
p <- p + geom_gate(c("CD4","CD8")) # short for geom_gate("CD8") + geom_gate("CD4")
p
```

By default, stats for all gate layers are added through empty `geom_stats` layer. 
```{r}
p + geom_stats() + labs_cyto("marker")
```

Note that we choose to only display marker on axis through `labs_cyto` layer here.

To add stats just for one specific gate, we can pass the gate name to `geom_gate`
```{r}
p + geom_stats("CD4")
```

stats type, background color and position are all adjustable.
```{r}
p + geom_stats("CD4", type = "count", size = 6,  color = "white", fill = "black", adjust = 0.3)
```

When 'subset' is not specified, it is at abstract status thus can not be visualized 
```{r error=T}
p <- ggcyto(gs, aes(x = CD4, y = CD8)) + geom_hex() + myPars
p
```

unless it is instantiated by the gate layer, i.e. lookup the gating tree for the parent node based on the given gates in `geom_gate`
```{r}
p + geom_gate(c("CD4", "CD8"))
```


Alternatively, we can choose to plot all children of one specified parent and projections
```{r}
p <- ggcyto(gs, aes(x = 38, y = DR), subset = "CD4") + geom_hex(bins = 64) + geom_gate() + geom_stats()
p
```


Or we can add gate layer to any arbitary node instead of its parent node
```{r}
ggcyto(gs, subset = "root", aes(x = CD4, y = CD8)) + geom_hex(bins = 64) + geom_gate("CD4") + myPars
```


Sometime it is helpful to display the axis label in raw scale by inverse transforming the axis without affecting the data
```{r}
p + axis_x_inverse_trans() + axis_y_inverse_trans()
#add filter (consistent with `margin` behavior in flowViz)
# ggcyto(gs, aes(x = CD4, y = CD8), subset = "3+", filter = marginalFilter)  + geom_hex(bins = 32, na.rm = T)
```

`ggcyto` is the semi-ggplot object since the data slot is NOT fully fortified to data.frame until it is printed/plotted.
```{r}
class(p)
class(p$data)
```

However it is still completely ggplot compatible in terms of adding layers and parameters.
We can convert it to a pure ggplot object
```{r}
p <- as.ggplot(p)

class(p)
class(p$data)
```




