---
title: 'ggcyto : Visualize `Cytometry` data with `ggplot`'
output:
  html_document:
    fig_height: 2
    fig_width: 2
    keep_md: yes
    toc: yes
    toc_float: true
vignette: >    
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Feature summary of ggcyto}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, fig.height= 3, fig.width= 5)
```

```{r}
library(ggcyto)
dataDir <- system.file("extdata",package="flowWorkspaceData")
```

## 1: suppoort `3` types of plot constructor 

* represent different levels of complexity and flexibility
* meet the needs of various plot applications
* suitable for users at different levels of coding skills.

### low level: `ggplot`

The overloaded `fority` methods empower `ggplot`  to work with all the major Cytometry data structures right away, which allows users to do all kinds of highly customized and versitled plots.

#### `GatingSet`
```{r}
gs <- load_gs(list.files(dataDir, pattern = "gs_manual",full = TRUE))
attr(gs, "subset") <- "CD3+"
ggplot(gs, aes(x = `<B710-A>`, y = `<R780-A>`)) + geom_hex(bins = 128) + scale_fill_gradientn(colours = gray.colors(9))
```

#### `flowSet/ncdfFlowSet/flowFrame` 
```{r}
fs <- getData(gs, "CD3+")
ggplot(fs, aes(x = `<B710-A>`)) + geom_density(fill = "blue", alpha= 0.5)
```

#### `gates`
```{r}
gates <- filterList(getGate(gs, "CD8"))
ggplot(gs, aes(x = `<B710-A>`, y = `<R780-A>`)) + geom_hex(bins = 128) + geom_polygon(data = gates, fill = "transparent", col = "purple")
```

### medium level: `ggcyto`

`ggcyto` constructor along with overloaded `+` operator encapsulate lots of details that might be tedious and intimidating for many users.

```{r}
ggcyto(gs, aes(x = CD4, y = CD8)) + geom_hex(bins = 128) + geom_gate("CD8")
```

It simplies the plotting by:
* add a default scale_fill_gradientn for you
* fuzzy-matching in `aes` by either detector or fluorochromes names
* determine the `parent` popoulation automatically
* exact and plot the gate object by simply referring to the `child` population name

### top level: `autoplot`
Inheriting the spirit from ggplot's `Quick plot`, it further simply the plotting job by hiding more details from users and taking more assumptions for the plot.

* when plotting `flowSet`, it determines `geom` type automatically by the number of `dim` supplied 
* for `GatingSet`, it further skip the need of `dim` by guessing it from the `children` gate

```{r}
#1d
autoplot(fs, "CD4")
#2d
autoplot(fs, "CD4", "CD8", bins = 64)

autoplot(gs, c("CD4", "CD8"), bins = 64)
```

## 2: in-line transformation 
It is done by different `scales` layers speically designed for `cytometry`
```{r}
data(GvHD)
fr <- GvHD[[1]]
p <- autoplot(fr, "FL1-H")
p #raw scale
p + scale_x_logicle() #flowCore logicle scale
p + scale_x_flowJo_fasinh() # flowJo fasinh
p + scale_x_flowJo_biexp() # flowJo biexponential

```

## 3: generic `geom_gate` layer 
It hides the complex details pf plotting different geometric shapes

```{r}
fr <- fs[[1]]
p <- autoplot(fr,"CD4", "CD8") + ggcyto_par_set(limits = "instrument")
#1d gate vertical
gate_1d_v <- openCyto::gate_mindensity(fr, "<B710-A>")
p + geom_gate(gate_1d_v)
#1d gate horizontal
gate_1d_h <- openCyto::gate_mindensity(fr, "<R780-A>")
p + geom_gate(gate_1d_h)
#2d rectangle gate
gate_rect <- rectangleGate("<B710-A>" = c(gate_1d_v@min, 4e3), "<R780-A>" = c(gate_1d_h@min, 4e3))
p + geom_gate(gate_rect)
#ellipsoid Gate
gate_ellip <- getGate(gs[[1]], "CD4")
class(gate_ellip)
p + geom_gate(gate_ellip)
```



## 4: `geom_stats` 
```{r}
p <- ggcyto(gs, aes(x = "CD4", y = "CD8"), subset = "CD3+") + geom_hex()
p + geom_gate("CD4") + geom_stats()
p + geom_gate("CD4") + geom_stats(type = "count") #display cell counts 
```


## 5: `axis_inverse_trans` 
It can display the `log` scaled data in the original value
```{r}
p # axis display the transformed values
p + axis_x_inverse_trans() # restore the x axis to the raw values
```

It currently only works with `GatingSet`.

## 6: auto limits
Optionally you can set limits by `instrument` or `data` range
```{r}
p <- p + ggcyto_par_set(limits = "instrument")
p
```

## 7: labs_cyto
You can choose between `marker` and `channel` names (or `both` by default)
```{r}
p + labs_cyto("markers")
```


## 8: `ggcyto_par_set` 
It aggregates the different settings in one layer
```{r}
#put all the customized settings in one layer
mySettings <- ggcyto_par_set(limits = "instrument"
                             , facet = facet_wrap("name")
                             , hex_fill = scale_fill_gradientn(colours = rev(RColorBrewer::brewer.pal(11, "Spectral")))
                            , lab = labs_cyto("marker")
                            )
# and use it repeatly in the plots later (similar to the `theme` concept)
p + mySettings

```

Currently we only support `4` settings, but will add more in future.

## 9: `as.ggplot` 
It allows user to convert `ggcyto` objects to pure `ggplot` objects for further the manipulating jobs that can not be done within `ggcyto` framework.
```{r}
class(p) # may not fully compatile with all the `ggplot` functions
p1 <- as.ggplot(p)
class(p1) # a pure ggplot object, thus can work with all the `ggplot` features
```

## 10: ggcyto_layout

Layout many gate plots on the same page

When plooting a `GatingHierarchy`, multiple cell populations with their asssociated gates can be plotted in different panels of the same plot.
```{r, fig.height = 4}
gh <- gs[[1]]
nodes <- getNodes(gh, path = "auto")[c(3:9, 14)]
nodes
p <- autoplot(gh, nodes, bins = 64)
class(p)
p
```

As you see, this generates a special `ggcyto_GatingLayout` object which is a container storing multiple `ggcyto` objects.
You can do more about the plot layout with the helper function `ggcyto_arrange`. 
For example, to arrange it as one-row gtable object
```{r}
gt <- ggcyto_arrange(p, nrow = 1)
class(gt)
plot(gt)
```

or even combine it with other `ggcyto_GatingLayout` objects(or any `gtable` objects) and print it on the sampe page
```{r}
p2 <- autoplot(getData(gh, "CD3+")[,5:8]) # some density plot
p2@arrange.main <- ""#clear the default title
gt2 <- ggcyto_arrange(p2, nrow = 1)

gt3 <- gridExtra::rbind.gtable(gt, gt2)
plot(gt3)

```

